The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-08-22 paper/paper.Rnw
Explored tracking and build some nice models in the end

<<default_conf>>=
  library(matrixStats)
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)
  library(ggthemes)

@@

results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

\begin{figure}
\begin{center}
<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps acros countries
diff_increase_fun <- function(df) {

@@

pooled_trendline <-
  complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(!is.na(country))

pooled_trendline %>%
  filter(type_test == "math") %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")
@
\caption{Evolution of the achievement gap for selected countries}
\end{center}
\end{figure}

@@

% Code chunks for the same plots as above for reading. Available upon request.
<<evolution_reading, eval = F>>=
# See below that reading graph is upon request. This is the one.

pooled_trendline %>%
  filter(type_test == "read") %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  coord_fixed() +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

perc_graph(top_bottom_perc, "read", "90/10 achievement gap", "Percentage change from 2000 to 2015")
@

@@

\section{Appendix}

<<>>=
results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))

# results_math <- read_rds("./data/delete.Rdata")
# results_read <- read_rds("./data/delete_read.Rdata")
# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")

>>>>

<<default_conf>>=
library(knitr)
library(arm)
library(saves)
library(haven)
library(PISA2000lite)
library(PISA2003lite)
library(PISA2006lite)
library(PISA2009lite)
library(PISA2012lite)
library(intsvy)
library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
library(countrycode) # For region variable
library(car)
library(SAScii)
library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
library(lme4)
library(modelr)
library(ggrepel)
library(xtable)
library(ggthemes)
library(matrixStats)
library(WDI)
library(tidyverse)

@@

# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))

# results_math_75 <- results_math
# results_read_75 <- results_read
# results_math_topmid_75 <- results_math_topmid
# results_read_topmid_75 <- results_read_topmid
# results_math_midbottom_75 <- results_math_midbottom
# results_read_midbottom_75 <- results_read_midbottom

results_math <- read_rds("./paper/data/delete.Rdata")
results_read <- read_rds("./paper/data/delete_read.Rdata")
results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
# US is missing for reading

@@

\begin{figure}
\begin{center}
<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
# 90/10 gaps across countries

df <- complete_data_topbottom

diff_increase_fun <- function(df) {

@@

p <- 
  complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "math", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  facet_wrap(~ country, ncol = 5) +
  geom_linerange(aes(ymin = 0, ymax = difference))

p + geom_line(data = complete_data_topbottom %>%
                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
                filter(!is.na(country)),
              stat = "smooth", method = "lm", aes(group = 1),
              formula = y ~ splines::ns(x, 2), size = 0.7,
              colour = "red") +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

@
\caption{Evolution of the achievement gap for selected countries}
\end{center}
\end{figure}

<<>>=
m <- complete_data_topbottom %>%
  filter(type_test == "math") %>%
  select(wave, country, difference) %>%
  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
  left_join(central_examination) %>%
  mutate(num_tracks = factor(tracks15y))

m %>%
  filter(country %in% countries) %>%
  split(.$country) %>%
  map(~ lm(difference ~ wave + ztrack, data = .x))

p + geom_smooth(data = complete_data_topbottom %>%
                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
                filter(!is.na(country)),
                method = "lm", se = FALSE,
                aes(group = 1), size = 0.7,
                colour = "red") +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")
@

@@

% Code chunks for the same plots as above for reading. Available upon request.
<<evolution_reading, eval = F>>=
# See below that reading graph is upon request. This is the one.
complete_data_topbottom %>%
  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
  filter(type_test == "read", !is.na(country)) %>%
  ggplot(aes(as.character(wave), difference)) +
  geom_point() +
  geom_linerange(aes(ymin = 0, ymax = difference))+
  geom_line(stat = "smooth", method = "lm", aes(group = 1),
            formula = y ~ splines::ns(x, 2), size = 0.7,
            colour = "red") +
  facet_wrap(~ country, ncol = 5) +
  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
  coord_fixed() +
  ggthemes::theme_few() +
  theme(panel.spacing = unit(1, "lines"),
        panel.grid.major.y = element_line(colour = "grey")) +
  ggtitle("Evolution of the 90/10 achievement gap")

  perc_graph(top_bottom_perc, "read", "90/10 achievement gap", "Percentage change from 2000 to 2015")

@@

<<eval = F>>=

# 2000
# SC03Q01 School public/private
# 1 Public
# 2 Private
# 7 N/A
# 8 M/R
# 9 Mis

# 2003
# SC03Q01 (8) Public or private
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Miss

# 2006
# SC02Q01 (8) Public or private
#  1 Public
#  2 Private
#  7 N/A
#  8 Invalid
#  9 Missing

# 2009
# SC02Q01
# 1 public
# 2 private

# 2012
# SC01Q01 Public or private Num
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing

# 2015
# SC013Q01TA
# 1 Public
# 2 Private
# 7 N/A
# 8 Invalid
# 9 Missing

# 2000: ST25Q01
# 1 <ISCED 2A>
# 2 <ISCED 2B>
# 3 <ISCED 2C>
# 4 <ISCED 3A>
# 5 <ISCED 3B>
# 6 <ISCED 3C>
# 7 N/A
# 8 M/R
# 9 Mis

# 2003: ISCEDO
# 2006: ISCEDO
# 2009: ISCEDO
# 2012: ISCEDO
# 2015: ISCEDO

# 1	General
# 2	Pre-Vocational
# 3	Vocational
# 4	Modular

library(readxl)

wd_extract <- function(var, name) {
  wdi_data <-
    WDI(indicator = var, start = 2000, end = 2015) %>%
    select(-1) %>%
    as_tibble() %>%
    mutate(year = as.character(year))

  names(wdi_data)[2] <- name
  wdi_data
}

pop_preprimary <- wd_extract("SP.PRE.TOTL.IN", "pop_preprimary")
teacher_pupil_ratio <- wd_extract("SE.PRE.ENRL.TC.ZS", "teacher_pupil_ratio")
perc_teacher_trained <- wd_extract("UIS.TRTP.0", "perc_teach")
preprimary_priv_enrol <- wd_extract("SE.PRE.PRIV.ZS", "preprimary_priv_enrol")

tracking_data <-
  readxl::read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
  map_if(is_double, round, 2) %>%
  as_tibble

difference_data <-
  complete_data_topbottom %>%
  filter(type_test == "math") %>%
  select(wave, country, difference) %>%
  rename(year = wave) %>%
  mutate(year = as.character(year))

library(inequalityintsvy)

central_examination <-
  tribble(
  ~ country, ~central_examination,
  "Australia", 1, 
  "Austria", 0, 
  "Belgium", 0,
  "Bulgaria", 1,
  "Canada", rbinom(1, 1, prob = 0.51),
  "Czech Republic", 1,
  "Denmark", 1,
  "Finland", 1,
  "France", 1,
  "Germany", 0,
  "United Kingdom", 1,
  "Greece", 0,
  "Hong Kong", 1,
  "Hungary", 1,
  "Iceland", 1,
  "Ireland", 1,
  "Israel", 1,
  "Italy", 1,
  "Japan", 1,
  "Korea", 1,
  "Latvia", 1,
  "Liechtenstein", 1,
  "Luxembourg", 1,
  "Netherlands", 1,
  "New Zealand", 1,
  "Norway", 1,
  "Poland", 1,
  "Portugal", 0,
  "Russia", 1,
  "Slovakia", 1,
  "Slovenia", 1,
  "Spain", 0,
  "Sweden", 0,
  "Switzerland", 0,
  "Turkey", 1,
  "United States", 1
  )

# The tracking variables explain very nicely the difference in test scores
# for all countries pooled, up to 50% R squared.

ready_data <-
  difference_data %>%
  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
  left_join(central_examination) %>%
  mutate(num_tracks = factor(tracks15y),
         age_selection = ifelse(selage >= 15, 1, 0))

mod_tracking <-
  ready_data %>%
  lm(difference ~
       num_tracks +
       age_selection +
       length +
       central_examination, data = .)

new_data <-
  data_grid(ready_data,
            selage, num_tracks, central_examination, .model = mod_tracking) %>%
  filter(complete.cases(.))

new_data$pred <- predict(mod_tracking, newdata = new_data)

new_data %>%
  ggplot(aes(selage, pred, colour = num_tracks)) +
  geom_point()

new_data %>%
  ggplot(aes(num_tracks, pred, fill = as.character(num_tracks))) +
  geom_col(position = "dodge") +
  facet_wrap(~ central_examination) +
  scale_fill_fivethirtyeight()

# What happened if you adjut the trend splines for the tracking data?
@

<<>>=

# One idea I had was to make a cumulative sum of the trend,
# so the cumulative inequality over time for each country
# and see how each tracking feature explains this cumulative
# pattern, and it does so very well, with each one
# separately (because I have only 30 observations) explaining
# about 22% each. Age selection, up to 30% alone.

ready_data_two <-
  difference_data %>%
  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
  left_join(central_examination) %>%
  mutate(num_tracks = factor(tracks15y),
         age_selection = ifelse(selage >= 15, 1, 0))

ready_data_two <-
  ready_data_two %>%
  group_by(country) %>%
  summarize(
    cum_diff = sum(difference, na.rm = TRUE),
    length = unique(length),
    num_tracks = unique(num_tracks),
    age_selection = unique(age_selection),
    central_examination = unique(central_examination),
    avg_track = unique(ztrack)
         )

mod_tracking <-
  ready_data_two %>%
  lm(cum_diff ~ age_selection, data = .)
@
